---
title: "BAE 204 - Lab11_Homework #7"
author: "First Lastname"
date: "due 3 April 2020 before midnight"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height=5, fig.width=9, fig.align="center", echo = FALSE)
```

# Exercise 1

Below, the hydrograph for the year 2008-2009 for the Cuyahoga River at Independence, OH is plotted. 

```{r , echo=TRUE, eval=TRUE}

data<-read.csv(file="https://raw.githubusercontent.com/francoisbirgand/BAE204-Qlab/master/Lin_1h_cuyahoga_all_2008-2009_SI_ref.csv",header = TRUE) #Reads file into table format

WSarea<-707*1.60934^2 #Area of watershed in km2
WS<-"Cuyahoga River at Independence, OH"

data<-as.data.frame(data)
data$datetime<-as.POSIXct(strptime(data$datetime, "%Y-%m-%d %H:%M:%S")) # transforms characters into date values understood by R
D<-data$datetime
Q<-data$Q   #Defines Q as the flow value (m3/s)
C<-data$NO3   #Defines C as the concentration value (mg N/L)

N=nrow(data)   #Sets N to the value equal to the number of total rows in the table

head(data) # This shows the first 5 rows of the data data.frame. It shows the names of each of the columns
names(data) # this shows a list of the names of the columns of data

```

# Exercise 2

Calculate the lowest, highest, mean, and median flow rates, as well as the 10th percentile, and the 90th percentile. Report them in a bullet list using embedded code.


- The mean flow is `r signif(mean(Q),digits=3)` m^3^/s
- the 10^th^percentile of flow is `r signif(quantile(Q, probs = 0.1),digits=3)` m^3^/s

# Exercise 3

Calculate and report the arithmetic average and medium concentrations for nitrate, TP, SRP, and TSS over the entire year of record.

The reference values for agricultural watersheds are, respectively, 5 mg N/L, 0.3 mg P/L, 0.03 mg P/L, and 20 mg TSS/L. Compare your results to these reference values.


```{r , echo=FALSE, eval=TRUE}
NO3<-data$NO3
TP<-data$TP
SRP<-data$SRP
```

- The mean concentration for nitrate is `r signif(mean(NO3),digits=3)` mg N/L

# Exercise 4

Plot the hydrograph and nitrate, TP, SRP, and TSS chemographs for the Cuyahoga River at Independence, Ohio, over the first two weeks of March 2009. Using text, report whether the chemographs exhibit a concentration or a dilution effect.

# Exercise 5

Plot the flow duration curve over the entire period of record using the log/log scale graph.
Report the total flow volume that occurred in 2%, 5%, and 10% of the time, corresponding to the highest flow.


# Exercise 6

Plot the load duration curve over the entire period of record using the log/log scale graph.
Report the total load that occurred in 2%, 5%, and 10% of the time, corresponding to the highest loads.

Compare these values to the ones you obtained for flow, and discuss whether or not that makes sense.

