---
title: "BAE 204 - Lab11_Homework #7"
author: "François Birgand"
date: "due 3 April 2020 before midnight"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height=5, fig.width=9, fig.align="center", echo = FALSE)
```

# Exercise 1

Below, the hydrograph for the year 2008-2009 for the Cuyahoga River at Independence, OH is plotted. 

```{r , echo=TRUE, eval=TRUE}

data<-read.csv(file="Lin_1h_cuyahoga_all_2008-2009_SI_ref.csv",header = TRUE) #Reads file into table format

WSarea<-707*1.60934^2 #Area of watershed in km2
WS<-"Flow Rate at Cuyahoga River at Independence, OH"


```

# Exercise 2

Calculating the lowest, highest, mean, median, as well as the 10th and 90th percentile of the flow rates yields:


- The mean flow was ```#r signif(mean(Q),digits=3)``` m^3^/s
- the 10^th^percentile of flow was ____ m^3^/s
- The mean flow rate was ___ m^3^/s.
- The median flow rate was ___ m^3^/s.
- The 10^th^ percentile of flow rate was ___  m^3^/s.
- The 90^th^ percentile of flow rate was ___  m^3^/s.


# Exercise 3

Calculating and comparing the arithmetic average and medium concentrations for nitrate, TP, SRP, and TSS over the entire year of record. As of note, the reference values for agricultural watersheds are, respectively, 5 mg N/L, 0.3 mg P/L, 0.03 mg P/L, and 20 mg TSS/L. Compare your results to these reference values.


```{r , echo=FALSE, eval=TRUE}
NO3<-data$NO3
TP<-data$TP
SRP<-data$SRP
```

- The mean concentration for nitrate was ___ mg N/L
- The median concentration for nitrate was ___ mg N/L.

And so on. Compare mean versus median and discuss the difference. And there almost the same, or not. If they were really different, think about what would is say about the occurrence of high values. Compare them to the reference values we have seen in class. Similarly discuss your findings for the other parameters.

# Exercise 4

Plot the hydrograph and nitrate, TP, SRP, and TSS chemographs for the Cuyahoga River at Independence, Ohio, over the first two weeks of March 2009. Using text, report whether the chemographs exhibit a concentration or a dilution effect.

# Exercise 5

Plot the flow duration curve over the entire period of record using the log/log scale graph.
Report the total flow volume that occurred in 2%, 5%, and 10% of the time, corresponding to the highest flow.


# Exercise 6

Plot the load duration curve over the entire period of record using the log/log scale graph.
Report the total load that occurred in 2%, 5%, and 10% of the time, corresponding to the highest loads.

Compare these values to the ones you obtained for flow, and discuss whether or not that makes sense.

