---
title: "Quick tutorial for linear regression"
author: "François Birgand"
date: "4/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Creating linear regression between y and x

Let us take these two variables x and y and plot them together

```{r }
x <- c(151, 174, 138, 186, 128, 136, 179, 163, 152, 131)
y <- c(63, 81, 56, 91, 47, 57, 76, 72, 62, 48)

plot(x,y)
```

Now, we want to find a linear model ```(lm)``` that best fits through the cloud of points such that in the end we will get a model $y = ax + b$, and we want to find the two coefficients $a$ and $b$.

for this we use the following expression where a variable named ```relation``` has all the attibutes associated with a linear regression of y as a function of x. In this form, it assumes that the intercept is not 0

```{r }
relation <- lm(y~x) 
```


### How to retrieve some of the values associated with the linear model
#### When intercept is not 0

```{r }
names(relation) # 

```

gives quite a few important parameters that are included in the object relation.


</br>



</br>

```{r }
summary(relation)
```

gives you a summary of important (same as above + additional) values compiled into a readable table. To obtain some of the values like the R&sup2; or other things.

</br>

</br>

```{r }
names(summary(relation))
```

gives you the names of all the variables that appear as a readable table when typing ```summary(relation)```. 

</br>

```{r }
relation$coefficients # is a vector that gives the coefficients of the regression 
relation$coefficients[1] # gives the intercept of the regression line or the coefficient b
relation$coefficients[2] # gives the slope of the regression line or the coefficient a
```

gives you the names of all the variables that appear as a readable table when typing ```summary(relation)```. 

</br>

You may thus also retrieve the coefficients of the regression using

```{r }
summary(relation)$coefficients # 
```

You now have also access to the standard error on the intercept (first row) and on the slope (second row), the t value  gives the results of the t-test associated with testing the significance of the parameter listed in the first column, and Pr(>|t|) gives you the p-value for that t-test.

</br>

```{r}
# You could thus retrieve the value for the slope using 
summary(relation)$coefficients[2,1] #second row of the first column 

# You could thus retrieve the value for the intercept using 
summary(relation)$coefficients[1,1] #first row of the first column 
```

So essentially ```relation$coefficients[1]``` is the same as ```summary(relation)$coefficients[1,1]``` and
 ```relation$coefficients[2]``` is the same as ```summary(relation)$coefficients[2,1]```.
 
So may use the values of the coefficients you have found to calculate anything you need with them.

</br>

#### When intercept is 0

You essentially use the same tools but the synthax for the regression becomes:

```{r }
relation_no_intercept <- lm(y~x+0) 
relation_no_intercept$coefficients # now only has one value for the slope
```

</br>

### How to plot the linear model

There are several ways to plot the variables. The first one is to use the function ```abline``` as below. That is very efficient.

```{r }
x <- c(151, 174, 138, 186, 128, 136, 179, 163, 152, 131)
y <- c(63, 81, 56, 91, 47, 57, 76, 72, 62, 48)


plot(x,y)
abline(relation,col='red')
abline(relation_no_intercept,col='green')

```

</br>

You may want to express directly the expression of the coefficients in the equation just like below.

</br>

```{r }

xlim=c(100,200);ylim=c(40,100) # important to define x and y scales now as you are adding a new plot and the automatic scale
# adjustment may not be the same for both curves
plot(x,y, xlim=xlim,ylim=ylim)

a <- relation$coefficients[2]
b <- relation$coefficients[1]
alpha <- relation_no_intercept$coefficients

par(new=TRUE) # just tells R to wait to plot an additional curve

plot(x,a*x+b, xlim=xlim,ylim=ylim, type='l', col = 'red',ylab='')

par(new=TRUE)
plot(x,alpha*x, xlim=xlim,ylim=ylim, type='l', col = 'green',ylab='')

```

</br>

But you may also use a totally different ```x``` in the plotting.

</br>

```{r }
xx <- seq(0,300,by=1) # xx is now a vector from 0 to 300 at increment equal to 1

xlim=c(100,200);ylim=c(40,100) # important to define x and y scales now as you are adding a new plot and the automatic scale
# adjustment may not be the same for both curves
plot(x,y, xlim=xlim,ylim=ylim)

a <- relation$coefficients[2]
b <- relation$coefficients[1]
alpha <- relation_no_intercept$coefficients

par(new=TRUE) # just tells R to wait to plot an additional curve

plot(xx,a*xx+b, xlim=xlim,ylim=ylim, type='l', col = 'red',ylab='')

par(new=TRUE)
plot(xx,alpha*xx, xlim=xlim,ylim=ylim, type='l', col = 'green',ylab='')

```
